---
title: "Domain Interest vs. Language Preference Statistics"
author: "Adarsh Nagar"
date: "2025-10-09"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction:

Today we will be exploring the relationship between the language comfortability/preference from the intake form and individual domain interest for 2025 Data
Science Capstone projects.



### Import the data
```{r, warning=FALSE}
library(tidyverse)

# retrieve class survey data
url <- 'https://raw.githubusercontent.com/pstat197/pstat197a/main/materials/labs/lab2-tidyverse/data/'

background <- paste(url, 'background-clean.csv', sep = '') %>%
  read_csv()

interest <- paste(url, 'interest-clean.csv', sep = '') %>%
  read_csv()
interest$lang[interest$response_id == 29] <- "Python"
interest$dom[interest$response_id == 29] <- "Technology;Software development;Product engineering"

metadata <- paste(url, 'survey-metadata.csv', sep = '') %>%
  read_csv()

```

### EDA & Satisfying Expectations

```{r}
library(ggplot2)
library(dplyr)

# Extract exact columns
consolidated_csv <- interest[3:4]
consolidated_df <- data.frame(consolidated_csv)

#Total Counts for each domain
expanded_df <- consolidated_csv |> 
  separate_rows(dom,sep =";") |>
  mutate(dom = trimws(dom))

# summarize to lang × dom counts 
plot_data <- expanded_df |>
  count(lang, dom, name = "n") |>
  mutate(lang = fct_reorder(lang, n, .fun = sum)) 

# Bar Plot (Frequency + Language)
stacked <- ggplot(plot_data, aes(y = n, x = lang, fill = dom)) +
  geom_col(position = position_dodge(width = 0.9)) +
  labs(title = "Plot 1: Domain Interest by Languge",
       x = "Language Preference", y = "Count", fill = "Domain") +
theme(legend.position = "right") + geom_hline(yintercept = 5,color="black") 
stacked

# Proportions Stacked Barplot
ggplot(expanded_df, aes(x = lang, fill = dom)) +
  geom_bar(position = "fill") +
  labs(title = "Plot 2: Proportion of Domain Interests by Language",
       x = "Language", y = "Proportion") +
  theme_minimal()

# List of Satisfying
tbl <- table(expanded_df$lang, expanded_df$dom)
chi <- suppressWarnings(chisq.test(tbl))
expected <- as.data.frame(as.table(chi$expected))
observed <- as.data.frame(tbl)


observed <- as.data.frame(tbl)
ggplot(observed, aes(x = Var1, y = Var2, fill = Freq)) +
  geom_tile(color = "grey70") +
  geom_text(aes(label = Freq), size = 3, color = "white") +
  scale_fill_gradient(low = "steelblue4", high = "lightblue") +
  labs(
    title = "Observed Contingency Table (Chi-Square Test Inputs)",
    fill = "Observed Count",
    x = "Language Preference",
    y = "Domain Interest"
  ) +
  theme_minimal()

```


# Statistical Testing
```{r}
# Monte Carlo Simulation Chi-Square Test
chi_sim <- chisq.test(tbl, simulate.p.value = TRUE, B = 10000)
print(chi_sim)
```


# Results & Findings

Examining the exhibits, the observed relationship between language preference and
domain interest data appears to be variable and slightly skewed. Specifically, observing
Plot 1, there appears to be a larger amount of interest and thus larger amounts 
of data within Python interest (specifically among Technology and Software 
Development), while the remaining two categories, still with a considerable 
density of information, appear greatly spread across all domains. 

This visual phenomena is further illustrated in Plot 2, where similar densities 
of domain interest in each specific language preference appears more similar 
by proportional comparison. Going forward, this insight will prove to beneficial
as a demonstration of visually-consistent variation among language groups.

As a preliminary measure to understanding the statistical eligibility for Chi-Square
Testing, Plot 3 presents a contingency table for Chi-Square Testing. Hoping to suffice
the requirement of $$ x > 5$$ for each cell of x observations between domains and languages,
this plot exposes the lack of satisfaction and poses an imminent threat to statistical accuracy.
Specifically, we realize 13 among the 45 cells fail to satisfy the Chi-Square Test.

Since the Asymptotic approximation of Chi-Square is not possible with such small
frequencies, we defer to the use of the Monte Carlo simulated Chi-Square Test to
generate a proportion of simulated statistics across 10,000 contingency tables.
This result has yielded a p-value of $$p = 0.6478$$, proving that our observed
statistic reinforces the result that there is no statistically significant 
association between programming language preference and domain of interest. Albeit
selected under such circumstances, potential drawbacks include a loss of randomness derived
in bootstrapping and computational intensity. While the size of the data is relatively small,
considering its observations contain the sole interests of the 2025 Capstone Cohort, perhaps
this sparseness could be combatted in the future with data aggregation across multiple
term cohorts.







