---
title: "Analysis of class surveys"
subtitle: "If you want a subtitle put it here"
author: "List names here"
date: last-modified
published-title: "Updated"
editor: visual
format: html
code-copy: true
execute:
  message: false
  warning: false
  echo: false
  cache: true
---

Use this as a template. Keep the headers and remove all other text. Overall, your report may be quite short. When it is complete, render and then push changes to your team repository.

```{r}
# load any other packages and read data here
library(tidyverse)
```

## Executive summary

Write a one-paragraph abstract summarizing what you did and your findings. It need not be comprehensive; try to highlight the most important or interesting outcomes.

## Data description

-   **Source:** Survey distributed to all students offered enrollment in **PSTAT197A, Fall 2025**.

-   **Sample Size:** 𝑛 = 60 responses.

-   **Sampling:** No random selection. Can be considered:

    -   A census of PSTAT197A enrollees, or

    -   A convenience sample of students interested in **data science/capstones**.

**Sample Characteristics:**

| Characteristic       | Count | Percent |
|----------------------|-------|---------|
| **Class Standing**   |       |         |
| Junior               | 9     | 15%     |
| Senior               | 51    | 85%     |
| **Gender Identity**  |       |         |
| Female               | 28    | 46.7%   |
| Male                 | 32    | 53.3%   |
| **Race / Ethnicity** |       |         |
| Asian                | 43    | 71.7%   |
| Caucasian            | 13    | 21.7%   |
| Multiracial          | 3     | 5%      |
| Prefer not to say    | 1     | 1.7%    |

**Consent for sharing:**

| Share background | Share project preferences | Count |
|------------------|---------------------------|-------|
| Yes              | Yes                       | 49    |
| Yes              | No                        | 2     |
| No               | Yes                       | 3     |
| No               | No                        | 6     |

**Measurements Collected:**

-   **Self-reported proficiency** in programming, math, and statistics: `beg`, `int`, `adv`.

-   **Comfort levels** (numeric scale 1–5) in programming (`prog.comf`), math (`math.comf`), statistics (`stat.comf`).

-   **Course history:** Counts of courses taken in departments like PSTAT, CS, LING, ECON.

-   **Past research experience**, project preferences, and demographic info.

**Notes on the dataset:**

-   Personal information, free-text responses, and students who did not consent were removed.

-   Some variables, like distinctions in research experiences, were simplified for analysis.

**Summary Statistics:**

| Variable  | Mean | Median | Min | Max |
|-----------|------|--------|-----|-----|
| prog.comf | 3.86 | 4      | 2   | 5   |
| math.comf | 4.04 | 4      | 3   | 5   |
| stat.comf | 4.04 | 4      | 2   | 5   |

-   Proficiency levels (numeric) roughly align with comfort ratings:

    -   `prog`: 2.9 (mean)

    -   `math`: 2.78

    -   `stat`: 2.83

## Questions of interest

Indicate the questions your analysis addresses. These should map one-to-one to your findings. Don't include questions you didn't answer or questions you started with and refined later. If you would rather frame them as goals or tasks rather than questions that is okay; just modify the header appropriately. However you frame what you've done, you may only have two or three items; that is fine. Provide an itemized or numbered list so that the reader can easily identify your objectives.

3.  Does a student’s preferred programming language (R, Python, or no preference) influence their self-reported proficiency and comfort in programming and math?

## Findings

Summarize your results. Don't try to explain every step you took; focus instead on providing the main data analytic outputs -- tables and figures -- and explaining clearly what they show. Clarify any important decisions you made in obtaining them. You can display codes if you like but it is not necessary.

The following analyses examine how students’ programming language preference (R, Python, or No preference) relates to their proficiency and comfort in programming and math.

**Comfort and Proficiency by Language**

We first converted students’ self-reported proficiency levels in programming and math into numeric scores (1 = beginner, 2 = intermediate, 3 = advanced) and combined them with their comfort scores. The table below shows the ANOVA results testing for differences in these scores by language preference:

```{r}
background <- read_csv("../data/background-clean.csv")
interest   <- read_csv("../data/interest-clean.csv")

# Inner join by response ID
data <- inner_join(background, interest, by = "response_id")

# Only select needed columns
data <- data %>%
  select(lang, prog.prof, prog.comf, math.prof, math.comf)

library(dplyr)

# Turn text proficiency levels into numbers
data <- data %>%
  mutate(
    prog.prof.num = recode(prog.prof, "beg" = 1, "int" = 2, "adv" = 3
    ),
    math.prof.num = recode(math.prof, "beg" = 1, "int" = 2, "adv" = 3
    )
  )

# Pivot longer
data_long <- data %>%
  pivot_longer(
    cols = c(prog.prof.num, prog.comf, math.prof.num, math.comf),
    names_to = "skill_type",
    values_to = "score"
  ) %>%
  mutate(
    skill_type = recode(skill_type,
                        "prog.prof.num" = "Programming Proficiency",
                        "prog.comf" = "Programming Comfort",
                        "math.prof.num" = "Math Proficiency",
                        "math.comf" = "Math Comfort")
  )

# Run ANOVA: Does comfort/proficiency differ by language?
anova_result <- aov(score ~ lang * skill_type, data = data_long)
summary(anova_result)
```

The ANOVA indicates that **there is a significant difference between skill types** (Programming vs Math, Proficiency vs Comfort), but **no significant differences between languages** or their interaction with skill type. In other words, students’ comfort and proficiency levels are fairly similar across those preferring R, Python, or having no preference.

**Visualization**

The boxplot below illustrates the distributions of programming and math comfort and proficiency scores by language preference:

```{r}
# Visualize mean differences
ggplot(data_long, aes(x = lang, y = score, fill = skill_type)) +
  geom_boxplot() +
  labs(
    title = "Comfort and Proficiency by Language Preference",
    x = "Language Preference",
    y = "Score (1–5 scale or recoded proficiency)",
    fill = "Skill Type"
  ) +
  theme_minimal()
```

From the plot, we see that **median scores are similar across language preferences** for both programming and math. Programming proficiency and comfort are slightly higher on average than math scores, consistent with the overall skill distribution in the class.

## Useful tips (remove this section)

Due to the visual editor, most common needs for authoring can be met using menu items in the RStudio IDE. For a comprehensive guide to Quarto, see the [documentation](https://quarto.org/docs/guide/); consult the documentation for anything you can't sort out in the IDE.

### Code chunks

Insert code chunks using the *Insert* drop-down menu or `ctrl + alt + I` . By default in this document, codes will not be shown but results will. If you want to change this behavior, add the chunk option `echo: true` as below.

```{r}
#| echo: true
my_df <- tibble(animal = c('snake', 'gecko'),
                threatening = c(TRUE, FALSE))

my_df
```

Code chunks will be executed in order when the document is rendered, so packages should be loaded at the very beginning of the document

### Rendering tables in markdown

Several functions exist for converting dataframes to markdown tables for nice display. A simple one from the `knitr` package is `knitr::kable()` .

```{r}
my_df %>% knitr::kable()
```

### Figures

To include any figures not generated by code chunks, use *Insert \> Figure / Image* or `![CAPTION](path/file.ext)` .

Most of the time code chunks producing plots will render fine. However, to adjust sizing or alignment, use the chunk options `fig-width` and `fig-height` . Captions can be added with `fig-cap` .

```{r}
#| fig-width: 4
#| fig-height: 3
#| fig-align: left
#| fig-cap: A terrible plot.

my_df %>% 
  ggplot(aes(x = animal, y = threatening)) + 
  geom_col()
```

### Links

To include any hyperlinks, use `[display text](url)` .
