---
title: "Analysis of class surveys"
subtitle: "If you want a subtitle put it here"
author: "List names here"
date: last-modified
published-title: "Updated"
editor: visual
format: html
code-copy: true
execute:
  message: false
  warning: false
  echo: false
  cache: true
---

Use this as a template. Keep the headers and remove all other text. Overall, your report may be quite short. When it is complete, render and then push changes to your team repository.

```{r}
# load any other packages and read data here
library(tidyverse)
library(ggplot2)
library(dplyr)
library(scales)

background <- read_csv('../data/background-clean.csv')
interest <- read_csv('../data/interest-clean.csv')

all_data <- background %>% inner_join(interest, by = join_by(response_id)) %>% rename(c(domain_yn = dom.x, domain_type = dom.y, research_or_industry = do_you_have_any_preference_regarding_working_on_an_industry_project_or_a_research_lab_project))

all_data$lang[all_data$response_id == 29] <- "Python"
```

## Executive summary

Write a one-paragraph abstract summarizing what you did and your findings. It need not be comprehensive; try to highlight the most important or interesting outcomes.

## Data description

Write a brief description of the dataset. Your description should cover how data were obtained, sample characteristics, and measurements taken. It does not need to be exhaustive, but sufficiently detailed to convey a clear high-level understanding of the dataset. You can utilize bullet points or tables if you feel it would help improve clarity.

## Questions of interest

Indicate the questions your analysis addresses. These should map one-to-one to your findings. Don't include questions you didn't answer or questions you started with and refined later. If you would rather frame them as goals or tasks rather than questions that is okay; just modify the header appropriately. However you frame what you've done, you may only have two or three items; that is fine. Provide an itemized or numbered list so that the reader can easily identify your objectives.

We sought to understand how students’ programming language preferences relate to their academic background, skills, and project interests.In particular, we examined whether language preference aligns with coursework patterns in different departments, students’ domain interests, project type preferences (industry vs. research), class level or standing, and overall academic background. Our analysis addressed the following questions:

1.  How does students’ academic background and preparation relate to their programming language preference?
2.  How do students’ programming language preferences influence their interests and coursework choices?

## Findings

Summarize your results. Don't try to explain every step you took; focus instead on providing the main data analytic outputs -- tables and figures -- and explaining clearly what they show. Clarify any important decisions you made in obtaining them. You can display codes if you like but it is not necessary.

The following chart displays the distribution of students that prefer each programming language in the class. As shown, a majority of students (around 60%) prefer Python, around 22% of students are not partial toward either language, and around 18% of the class prefers R.

```{r}
lang_summary <- all_data %>%
  count(lang) %>%
  mutate(prop = n / sum(n)) %>%
  arrange(desc(prop))

## Bar Chart Of Language Preference w/ Percents

ggplot(lang_summary, aes(x = reorder(lang, -prop), y = prop, fill = lang)) +
  geom_col() +
  geom_text(aes(label = scales::percent(prop, accuracy = 0.1)), 
            vjust = -0.5, size = 4) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Proportions of Language Preference",
    x = "Preferred Language",
    y = "Percentage"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

The bar chart below illustrates the relationship between students’ project type preferences and their preferred programming language. The figure shows that a majority of those who prefer Python (79%) prefer to work on an industry-sponsored project, whereas a majority of students that prefer R either do not prefer one type of project over another. This may be due to the fact that Python is more prevalent in the industry, so students who prefer Python are more likely to seek industry-sponsored projects rather that research-based projects, which tend to use R.

```{r}
plot_data <- all_data %>%
  count(lang, type) %>%
  group_by(lang) %>%
  mutate(prop = n / sum(n))

# Distribution of those who prefer each type of Project
# By Language Preference
ggplot(plot_data, aes(x = lang, y = prop, fill = type)) +
  geom_col(position = "fill") +
  geom_text(
    aes(label = percent(prop, accuracy = 1)),
    position = position_fill(vjust = 0.5),
    size = 4
  ) +
  labs(
    title = "Project Type Preference by Language Preference",
    x = "Preferred Language",
    y = "Proportion",
    fill = "Project Type Preference"
  ) +
  scale_y_continuous(labels = percent) +
  theme_minimal()

```

The following chart shows that programming language preference is associated with coursework patterns. Students preferring R have taken more PSTAT courses (5.22 average) compared to Python-preferring students (4.55), showing more PSTAT coursework and experience is correlated with R preference. Conversely, Python-preferring students have completed more CS courses (1.52 vs. 1.11), aligning with Python's prominence in computer science. ECON and LING courses show less variation across language preferences, suggesting they don't heavily influence language choice. Overall, the data indicates that students' language preferences reflect their disciplinary preparation.

```{r}
# course groups from list
course_cols <- c("PSTAT100", "PSTAT115", "PSTAT120", "PSTAT122", "PSTAT126", 
                 "PSTAT131", "PSTAT160", "PSTAT174", "CS9", "CS16", "LING104", 
                 "LING110", "LING111", "CS130", "CS165", "ECON145", "PSTAT127", 
                 "PSTAT134", "CS5")

pstat_courses <- c("PSTAT100", "PSTAT115", "PSTAT120", "PSTAT122", "PSTAT126", 
                   "PSTAT131", "PSTAT160", "PSTAT174", "PSTAT127", "PSTAT134")
cs_courses <- c("CS9", "CS16", "CS130", "CS165", "CS5")
ling_courses <- c("LING104", "LING110", "LING111")
econ_courses <- c("ECON145")

# average number of courses taken by language preference
# more CS courses -> Python
# more PSTAT, Ling --> R

course_averages <- all_data %>%
  filter(lang %in% c("R", "Python", "No preference")) %>%
  mutate(
    pstat_total = rowSums(select(., all_of(pstat_courses)), na.rm = TRUE),
    cs_total = rowSums(select(., all_of(cs_courses)), na.rm = TRUE),
    ling_total = rowSums(select(., all_of(ling_courses)), na.rm = TRUE),
    econ_total = rowSums(select(., all_of(econ_courses)), na.rm = TRUE)
  ) %>%
  group_by(lang) %>%
  summarise(
    n = n(),
    avg_pstat = mean(pstat_total, na.rm = TRUE),
    avg_cs = mean(cs_total, na.rm = TRUE),
    avg_ling = mean(ling_total, na.rm = TRUE),
    avg_econ = mean(econ_total, na.rm = TRUE),
    avg_total = mean(pstat_total + cs_total + ling_total + econ_total, na.rm = TRUE)
  )

course_avg_long <- course_averages %>%
  select(lang, avg_pstat, avg_cs, avg_ling, avg_econ) %>%
  pivot_longer(cols = starts_with("avg_"), 
               names_to = "course_type", 
               values_to = "avg_courses") %>%
  mutate(course_type = case_when(
    course_type == "avg_pstat" ~ "PSTAT",
    course_type == "avg_cs" ~ "CS",
    course_type == "avg_ling" ~ "LING",
    course_type == "avg_econ" ~ "ECON"
  ))

ggplot(course_avg_long, aes(x = course_type, y = avg_courses, fill = lang)) +
  geom_col(position = "dodge") +
  geom_text(
    aes(label = round(avg_courses, 2)),
    position = position_dodge(width = 0.9),
    vjust = -0.5,
    size = 3.5
  ) +
  labs(
    title = "Average Number of Courses Taken by Language Preference",
    x = "Course Department",
    y = "Average Number of Courses",
    fill = "Language Preference"
  ) +
  theme_minimal() +
  theme(legend.position = "top")
```

``` R
{r}
```

## Useful tips (remove this section)

Due to the visual editor, most common needs for authoring can be met using menu items in the RStudio IDE. For a comprehensive guide to Quarto, see the [documentation](https://quarto.org/docs/guide/); consult the documentation for anything you can't sort out in the IDE.

### Code chunks

Insert code chunks using the *Insert* drop-down menu or `ctrl + alt + I` . By default in this document, codes will not be shown but results will. If you want to change this behavior, add the chunk option `echo: true` as below.

```{r}
#| echo: true
my_df <- tibble(animal = c('snake', 'gecko'),
                threatening = c(TRUE, FALSE))

my_df
```

Code chunks will be executed in order when the document is rendered, so packages should be loaded at the very beginning of the document

### Rendering tables in markdown

Several functions exist for converting dataframes to markdown tables for nice display. A simple one from the `knitr` package is `knitr::kable()` .

```{r}
my_df %>% knitr::kable()
```

### Figures

To include any figures not generated by code chunks, use *Insert \> Figure / Image* or `![CAPTION](path/file.ext)` .

Most of the time code chunks producing plots will render fine. However, to adjust sizing or alignment, use the chunk options `fig-width` and `fig-height` . Captions can be added with `fig-cap` .

```{r}
#| fig-width: 4
#| fig-height: 3
#| fig-align: left
#| fig-cap: A terrible plot.

my_df %>% 
  ggplot(aes(x = animal, y = threatening)) + 
  geom_col()
```

### Links

To include any hyperlinks, use `[display text](url)` .
